---
title: "tidy-model"
author: "Jissel Maldonado"
date: "2025-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('tidyverse') ; library('modeldata')
```

## Introduction

This R Markdown document demonstrates my abilities to use models for data analysis using a dataset collected on crickets.

```{r load data, include=FALSE}
data(crickets, package = "modeldata")
names(crickets)
```

## What is that sound? 

In this report, we examine what factors predict a cricket's chirp rate. 

```{r summary, echo = FALSE}
summary(crickets)
ggplot(crickets, aes(x = rate)) + 
  geom_histogram(bins = 8) + 
  ggtitle("Distribution of the chirp rate of crickets") + 
  xlab('Chirp rate (per min.)')
```

Summary:

Number of observations: 31 Number of species: 2 Temperature range: min of 17.20, max of 30.40 Mean rate of chirping: 72.89

## Temperature affects chirp rate

```{r temp, echo = FALSE}
ggplot(crickets, aes(x = temp, y = rate)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate") + 
  ylab('Chirp rate (per min.)') + 
  xlab('Temperature (Celcius)')

temp_lm <- lm(rate ~ temp, crickets)

summary.lm(temp_lm)
```

Based on a scatter plot of temperature and chirping and a correlation test, it seems that as temperature increases one degree, the rate of chirping increases about 4.2 chirps per minute.

## Species-specific effects of temperature on chirping

```{r species, echo = FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) + 
  geom_point() + 
  geom_smooth(method = 'lm') + 
  ggtitle("Plot of temperature and chirp rate for two species of crickets") + 
  ylab('Chirp rate (per min.)') + 
  xlab('Temperature (Celcius)')

species_lm <- lm(rate ~ temp + species, crickets)

summary.lm(species_lm)
```

Based on the scatter plot of temperature, chirping, species, and a correlation test, it seems that as temperature increases one degree, the rate of chirping increases about 3.6 chirps per minute. O. exclamationis has a chirp rate 10 chirps per minute higher than O. niveus. Additionally, the model accounts for about 99% of the variance in the data. 

## Interactions

```{r species histogram, echo = FALSE}
ggplot(crickets, aes(x = rate, fill = species)) + 
  geom_histogram(position = 'identity', alpha = 0.7, bins = 8) + 
  ggtitle("Distribution of the chirp rate of crickets") + 
  xlab('Chirp rate (per min.)')
```

The O. exclamationis and O. niveus species occupy different temperature zones, which potentially impacts chirp rate.

```{r interactions, echo = FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) + 
  geom_point() + 
  geom_smooth(method = 'lm') + 
  ggtitle("Plot of temperature and chirp rate for two species of crickets") + 
  ylab('Chirp rate (per min.)') + 
  xlab('Temperature (Celcius)')

species_x_temp_lm <- lm(rate ~ temp + species + temp:species, crickets)
summary.lm(species_x_temp_lm)

anova(species_lm, species_x_temp_lm)
```

The species-specific model and species-temperature interaction model are not statistically significantly different with a p-value greater than 0.05. Therefore, we can rely on the species-specific model for prediction about chirp rates.

## Visualize and model diamond worth using price, carat, and cut 
```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
data("diamonds")
```

We will now examine what factors determine a diamond's worth. 

```{r diamondSummary, echo = FALSE}
summary(diamonds)
ggplot(diamonds, aes(x = price)) + 
  geom_histogram(bins = 8) + 
  ggtitle("Distribution of the price range of diamonds") + 
  xlab('diamond price')
```

Summary: 
5 cuts, minimum carat - 0.2 and max carat - 5.01, mean price of 3933

## carat affects diamond price

```{r carat, echo = FALSE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of carat(s) and diamond price") +
  ylab('Diamond price') +
  xlab('Carat(s)')
```

Based on the scatter plot, as carats increase, diamond price increases. 